% To generate a PNG file, you may have to compile this with
% xelatex --shell-escape <filename>.
\documentclass[12pt,varwidth=15.835cm,convert={density=96},border=1]{standalone}

% Load font stuff for XeTeX, including Unicode support
\usepackage{fontspec}
\usepackage{xunicode}
\usepackage{xltxtra}

% Set main fonts
\newfontfamily{\Tagati}[
    Scale=1.2,
    BoldFont={* Italic},
]{Tagati Book G}

\pagestyle{empty}

% Line spacing
\usepackage{setspace}
\newenvironment{ayeri}{
    \doublespacing
    \Tagati
}{
    \par
}

\begin{document}

%% MAIN PART %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{ayeri}
kmyonF pMdhye\_aNF/henF mino – minrY mirneri sitNF/tonF pMdhaaNF/henF minrY.

\hspace{0.25em} EnYrENF Atauy kaarYo nNy pMdhn ObFlonFsFki.
silF\-visye sris EnFvnNF, ANF mN miry AyonF yen tYaan/tYaansF lyeri khni, seri 
    gnFvyaasF pMdhy tonF, nj ANF nrisye AyonF\-ymF yen, ANF miNF sjliNojye 
    mitnYmF nNy kmo kjvo yaaj. ENF mN yomaarnF Ed/mineye lug bhisY kj, nj toNF 
    vksF tenF puleNeri, sitNF/toNF/nmoj AyonNF nj EnFvnNF, naarY 
    nsimyye\_aNF/henF pMdhn nj nNaanen tonF njnj. 
ANF myyo nYaanF/henF nNy, miNF tenubisojrej, mitMtoNF kdnY.
ANF ENYonF vihYmF miromaanYe\_asF kejnmF si s lMtYonF kdnY Apineri koMdNy, 
    nsimyye\_asF pMdhn nj nNaanen ObFlonFsFki.
ANF srojye EnFvnF sNlsF yen, ANF mN yomojy AyonF rNY tonF lug bhisY kj.
s senYonF gnYe nNy/henF – ANF rnYe gnF\-vy ANFli kjvo lomaayy vismF nj ANF 
    thnYe ledoymF, ymF mY blNYeNF pinYnF ynolej gumo hiro ye – ANF sry ErFsy
    bhisY sris pidimY trik sirutjyaanen – ANF nristonF lomaay risNF nj lMty
    vptnsF tonF.
\end{ayeri}

%% END %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}
