\documentclass[12pt,paper=letter]{scrartcl}

% Handle language and quotation marks
\usepackage[ngerman,english]{babel}
\usepackage{csquotes} % Put quotations in \q{}!
\SetBlockEnvironment{quotation}
\renewcommand*{\mkccitation}[1]{ (#1)}

% Set all margins to 2.54 cm
\usepackage[margin=1in]{geometry}
\widowpenalty10000 % Avoid widows like the plague!
\clubpenalty10000 % Avoid orphans like the plage, too!

% Make multiple columns available in single-column document
\usepackage{multicol}

% Make text colors and color names available
\usepackage[xetex]{xcolor}

% Load font stuff for XeTeX, including Unicode support
\usepackage{fontspec}
\usepackage{xunicode}
\usepackage{xltxtra}

% Set main fonts
\newfontfamily{\Tagati}[
    Scale=1.1,
    BoldFont={* Italic},
    HyphenChar=·,
]{Tagati Book G}

\setmainfont[
    Ligatures=TeX,
    Numbers=Lowercase,
]{Junicode}

\setsansfont[
    Ligatures=TeX,
    Numbers=Lowercase,
    Scale=MatchUppercase,
    BoldFont={Open Sans Condensed Bold},
]{Open Sans Condensed Light}

% Load BibLaTeX (using Biber), configure citation styles
% Uses <https://github.com/semprag/biblatex-sp-unified>
\usepackage[
    bibstyle=biblatex-sp-unified,
    citestyle=authoryear-comp,
    maxnames=2,
    backend=biber,
    safeinputenc,
]{biblatex}

% To make \textcite look like "Doe (2014: 213)"
\renewcommand*{\postnotedelim}{\addcolon\addspace}
\DeclareFieldFormat{postnote}{#1}
\DeclareFieldFormat{multipostnote}{#1}

% Enable generating files from this .tex file (for the bibliography) 
\usepackage{filecontents}

% Date etc.
\usepackage{datetime}
\renewcommand{\dateseparator}{/}

% Clickable links in footnotes, TOC, etc.
\usepackage[
    xetex,
    bookmarks=true,
    colorlinks=false,
    linktoc=section,
    hidelinks,
]{hyperref}

\usepackage{url}
\urlstyle{rm}

% Ability to include graphics and dealing with footnotes in descriptions
\usepackage{graphicx}
\usepackage[font={small,sf},labelfont={small,sf},format=plain]{caption}
\usepackage{subcaption}
\usepackage{wrapfig}
\setlength{\columnsep}{2\baselineskip}

% General headers and footers
\usepackage{fancyhdr}
\pagestyle{fancy}

\fancyhead[L]{} % empty
\fancyhead[C]{} % empty
\fancyhead[R]{\thepage}

\fancyfoot[L]{} % empty
\fancyfoot[C]{} % empty
\fancyfoot[R]{} % empty

\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

% First page headers and footers are different
\fancypagestyle{firstpage}{
    \fancyhead[L]{\sffamily \footnotesize \textbf{Benung. The Ayeri Language Resource}}
    \fancyhead[C]{} % empty
    \fancyhead[R]{\sffamily \footnotesize Carsten Becker · \yyyymmdddate\today}
    
    \fancyfoot[L]{} % empty
    \fancyfoot[C]{} % empty
    \fancyfoot[R]{\sffamily \footnotesize 
	\href{http://benung.nfshost.com}{http://benung.nfshost.com} · 
	\href{https://github.com/carbeck/benung-pdfs}{github:carbeck/benung-pdfs} · 
	\href{https://creativecommons.org/licenses/by-sa/4.0/}{CC~BY-SA~4.0}
    }
    
    \renewcommand{\headrulewidth}{0.5pt}
    \setlength\footskip{0.5in}
}

\usepackage{ifthen}
\ifthenelse{\value{page}=1}{\thispagestyle{firstpage}{\pagestyle{fancy}}}

% Line spacing
\usepackage{setspace}
\onehalfspacing
%\linespread{1.5}

% Rotate big tables
\usepackage{rotating}

% Line numbering in verse environment
\usepackage{lineno}

% Formatting of glosses
\usepackage{gb4e}
\let\eachwordone=\itshape
%\def\cgdepthstrut{\vrule height 0pt depth 2ex width 0pt}
%\def\eachwordone{\cgdepthstrut\itshape}
%\pretocmd{\glt}{\vspace{0.75ex}}{}{}
%\AtBeginEnvironment{exe}{\smaller}

% Nicer footnotes
\usepackage[bottom]{footmisc}
\deffootnote{0em}{1.5em}{\textsuperscript\thefootnotemark\enskip}
\renewcommand{\footnoterule}{\rule{0pt}{0pt}{\vspace*{-0pt}}}
\setlength{\footnotesep}{1.5em}

% Smaller font in block quotes
\usepackage{relsize,etoolbox}
\AtBeginEnvironment{quote}{\noindent\smaller}
\AtBeginEnvironment{quotation}{\smaller}

% Macros
\newcommand{\fw}[1]{\textit{#1}} % Foreign Word
\newcommand{\tit}[1]{\textit{#1}} % Title of a work
\newcommand{\q}[1]{\enquote{#1}} % Context-aware quotation
\newcommand{\qq}[1]{\enquote*{#1}} % Explicit sublevel quotation
\newcommand{\tsup}[1]{\textsuperscript{#1}} % Superscript
\newcommand{\markyellow}[1]{\colorbox{yellow}{#1}} % Yellow highlighter

\newcommand{\divider}{\vspace{0.5\baselineskip} \centerline{* * *} \vspace{0.5\baselineskip}}

\newcommand{\mor}[1]{\textsc{\lowercase{#1}}}
\newcommand{\ayr}[1]{{\Tagati #1}}

\newenvironment{ayeri}{
    %\hyphenpenalty=10000
    %\hbadness=10000
    \doublespacing
    \begin{multicols}{2}
    \Tagati
}{
    \end{multicols} \par
}

\newenvironment{mytitle}{
    \begin{center}
	\Large
	\sffamily\bfseries
	~\\
}{
    \end{center}
}

% blah
\usepackage{lipsum}

%% META INFORMATION IN PDF FILE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypersetup{%
    pdfinfo={%
	Title={Translation Challenge: The Beginning of Tolstoy's "Anna Karenina"},
	Author={Carsten Becker}
    }
}

%% BIBLIOGRAPHY DATABASE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{filecontents*}{\jobname.bib}
@article{becker,
    author = {Carsten Becker},
    title = {Flicking Switches: {Ayeri} and the {Austronesian} Alignment},
    journaltitle = {Benung. The Ayeri Language Resource},
    date = {2012-06-27},
    urldate = {2014-10-26},
    url = {http://benung.nfshost.com/archives/2444},
}

@incollection{schachter,
    author = {Paul Schachter},
    title = {The Subject in {Philippine} Languages: Topic, Actor, Actor-Topic, or None of the Above?},
    booktitle = {{Subject and Topic}},
    editor = {Charles N. Li},
    location = {New York},
    publisher = {Academic P},
    date = {1976},
    pages = {493--518},
}

@online{tolstoy,
    author = {Leo Tolstoy},
    title = {Anna Karenina},
    translator = {Constance Garnett},
    editor = {David Brannan and David Widger and Andrew Sly},
    organization = {Project Gutenberg},
    date = {2013-02-22},
    origdate = {1878},
    urldate = {2014-10-26},
    url = {http://www.gutenberg.org/ebooks/1399},
}

@online{universalsarchive,
    title = {The Universals Archive},
    editor = {Frans Plank and Thomas Mayer and Tatsiana Mayorava and Elena Filimonova},
    organization = {U Konstanz},
    date = {2009},
    urldate = {2014-10-26},
    url = {http://typo.uni-konstanz.de/archive/intro},
}
\end{filecontents*}
\addbibresource{\jobname.bib}

%% END OF PREAMBLE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% MAIN PART %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{mytitle}
    Translation Challenge: The Beginning of Tolstoy's \q{\citetitle{tolstoy}}
\end{mytitle}

\section{Text in English}

The text to be translated in this Translation Challenge is the initial passage
of \citeauthor{tolstoy}'s 1878 novel \citetitle{tolstoy}.\footnote{Hat tip to 
Steven Lytle for suggesting it.} The Ayeri translation here follows the English 
one by Constance Garnett (1901), which can be found on Project Gutenberg.

\blockcquote{tolstoy}{
\noindent Happy families are all alike; every unhappy family is unhappy in its 
own way.

Everything was in confusion in the Oblonskys’ house. The wife had discovered 
that the husband was carrying on an intrigue with a French girl, who had been a 
governess in their family, and she had announced to her husband that she could 
not go on living in the same house with him. This position of affairs had now 
lasted three days, and not only the husband and wife themselves, but all the 
members of their family and household, were painfully conscious of it. Every 
person in the house felt that there was no sense in their living together, and 
that the stray people brought together by chance in any inn had more in common 
with one another than they, the members of the family and household of the 
Oblonskys. The wife did not leave her own room, the husband had not been at 
home for three days. The children ran wild all over the house; the English 
governess quarreled with the housekeeper, and wrote to a friend asking her to 
look out for a new situation for her; the man-cook had walked off the day 
before just at dinner time; the kitchen-maid, and the coachman had given warning.
}

\section{Text in Ayeri}
\begin{ayeri}
\noindent kmyonF pMdhye\_aNF/henF mino – minrY mirneri sitNF/tonF pMdhaaNF/henF minrY.

EnYrENF Atauy kaarYo nNy pMdhn ObFlonFsFki.
silF\-visye sris EnFvnNF, ANF mN miry AyonF yen tyaan/tyaansF lyeri khni, seri 
gnFvyaasF pMdhy tonF, nj ANF nrisye Ayo\-nYmF yen, ANF miNF sjliNojye mitnYmF 
nNy kmo kjvo yaaj. ENF mN yomaarnF Ed/mineye lug bhisY kj, nj toNF vksF tenF 
puleNeri, sitNF/toNF/nmoj AyonNF nj EnFvnNF, naarY nsimyye\_aNF/henF pMdhn nj 
nNaanen tonF njnj. 
ANF myyo nYaanF/henF nNy, miNF tenubisojrej, mitMtoNF kdnY.
ANF ENYonF vihYmF miromaanYe\_asF kejnmF sisaa ANF lMtYo kdnY ApinF koMdNy, nsimyye\_asF pMdhn nj nNaanen ObFlonFsFki.
ANF srojye EnFvnF sNlsF yen, ANF mN yomojy AyonF rNY tonF lug bhisY kj.

--- 
\end{ayeri}

\noindent
\tsup{1}~Kamayon pandahajang-hen mino; minarya miraneri sitang-ton 
    pandahāng-hen minarya.

\tsup{2}~Enyareng atauya kāryo nangaya pandahana Oblonski.
\tsup{3}~Silvisaye sarisa envanang, ang manga miraya ayon yena cān-cānas layeri 
    Kahani, seri ganvayās pandahaya ton, nay ang narisaye ayonyam yena, ang 
    ming saylingoyye mitanyam nangaya kamo kayvo yāy.
\tsup{4}~Eng manga yomāran eda-mineye luga bahisya kay, nay tong vakas ten 
    pulengeri, sitang-tong-namoy ayonang nay envanang, nārya nasimayajang-hen 
    pandahana nay nangānena ton naynay.
\tsup{5}~Ang mayayo nyān-hen nangaya, ming tenubisoyrey, mitantong kadanya.
\tsup{6a}~Ang engyon vihyam miromānjas keynam sisā ang lanco kadanya apin 
    kondangaya, nasimayajas pandahana nay nangānena Oblonski.
\tsup{6b}~Ang saroyye envan sangalas yena, ang manga yomoyya ayon rangya ton 
    luga bahisya kay.

...

\section{Interlinear Breakdown}

\begin{exe} % SENTENCE 1
    \ex \label{ex:1}
    \gll Kama -yon pandaha -ye -ang =hen mino; mino -arya miran -eri sitang- ton 
	pandaha -ang =hen mino -arya. \\
    be.like \mor{-3PL.N} family \mor{-PL} \mor{-A} =all happy; happy \mor{-NEG} 
	way \mor{-INS} self- \mor{3PL.N.GEN} family \mor{-A} =every happy
	\mor{-NEG}. \\
    \glt \q{All happy families are alike; every unhappy family is unhappy 
	in its own way.}
\end{exe}

I was looking for a way to express \qq{alike} here, first trying to go with
\ayr{kmo} \fw{kamo} \qq{equal, same} and then realizing that it might in fact
be better to use the related verb \ayr{km/} \fw{kama-} \qq{be as … as …}
intransitively to mean \qq{be alike}, i.e. expressing similarity rather than 
equality.

In an earlier article on the blog, I wrote about \mor{N ADJ} order in Ayeri that
\textcquote{becker}{the predication in equative sentences seems to be 
interpreted in the way of a transitive sentence, although it is lacking an 
overt predicate}: in Ayeri, adjectives in predicative sentences come after
the subject NP, as though they were objects. This, in fact, goes counter to 
Universal 145 in \citetitle{universalsarchive} \autocite{universalsarchive},
since \citeauthor{universalsarchive} predict predicative adjectives in VSO 
languages to be treated like (or at least similar to, I suppose) verbs. I 
decided to break my rule and follow the universal in this case for stylistic 
reasons, since I thought that \fw{pandahāng-hen minarya minarya miraneri 
sitang-ton} does not flow as well as with the constituents reversed.

\divider

\begin{exe} % SENTENCE 2
    \ex
    \gll Enya -reng atau -ya kāryo nanga -ya pandaha -na Oblonski. \\
    everything \mor{-A.INAN} chaos \mor{-LOC} big house \mor{-LOC} family 
	\mor{-GEN} Oblonski. \\
    \glt \q{Everything was in a big chaos in the Oblonski family's house.}
\end{exe}

This sentence shows Ayeri's normal constituent order again, as compared to 
(\ref{ex:1}), the predication being the NP \q{atauya kāryo}. Tagalog,
which I have been returning to as an inspiration for Ayeri, seems to place 
not only predicative adjectives, but also predicative NPs before actors, however:

\begin{exe}
    \exi{(i)} Tagalog \autocite[499]{schachter}:
    \begin{xlist}
	\ex
	\gll Abogado ang lalaki. \\
	lawyer \mor{T-} man \\
	\glt \q{The man is a lawyer.}
	
	\ex
	\gll Matalino ang lalaki. \\
	intelligent \mor{T-} man \\
	\glt \q{The man is intelligent.}
    \end{xlist}
\end{exe}

Since I could not find out anything about the etymology of the name Oblonsky,
I left it as it is and respelled it with the graphemes available in Ayeri 
transcription.

\divider

\begin{exe} % SENTENCE 3
    \ex
    \begin{xlist}
	\ex \gll Silvisa -ye sarisa envan -ang, ang manga mira -ya ayon {} yena 
	    cān \textasciitilde{}cān -as lay -eri Kahani, si -eri ganvaya -as 
	    pandaha -ya ton, … \\
	Discover \mor{-3SG.F} previously wife \mor{-A}, \mor{AT} \mor{PROG} do
	    \mor{-3SG.M} husband \mor{(.T)} \mor{3SG.F.GEN} love 
	    \mor{\textasciitilde{}DIM} \mor{-P} girl \mor{INS} French, 
	    \mor{REL} \mor{-INS} governess \mor{-P} family \mor{-LOC} 
	    \mor{3PL.N.GEN}, … \\
	\glt \q{The wife had previously discovered that her husband
	    had been having an affair with a French girl, who had been a
	    governess in their family, …}
	
	\ex \gll …, nay ang narisa -ye {} ayon -yam yena, ang ming sayling -oy 
	    -ye {} mitan -yam nanga -ya kamo kayvo yāy. \\
	…, and \mor{AT} announce \mor{-3SG.F} \mor{(.T)} husband \mor{-DAT}
	    \mor{3SG.F.GEN}, \mor{AT} can continue \mor{-NEG} \mor{-3SG.F} 
	    \mor{(.T)} live \mor{-PTCP} house \mor{-LOC} same with 
	    \mor{3SG.M.LOC}. \\
	\glt \q{…, and had announced to her husband that she could not 
	    continue living in the same house with him.}
    \end{xlist}
\end{exe}

This is the first section that necessitated coming up with new vocabulary:
\ayr{tYaanF/tYaanF} \fw{cān-cān} \qq{(love) affair} as a diminutive of 
\ayr{tYaanF} \fw{cān} \qq{love} (a little like German \fw{Liebelei} and 
\fw{Techtelmechtel}); \ayr{gnFvy} \fw{ganvaya} \qq{governess} from a 
combination of \ayr{gnF} \fw{gan} \qq{child} and the feminine occupational 
derivative suffix \ayr{/vy} \fw{-vaya}; and \ayr{nris/} \fw{narisa-} 
\qq{announce} as a causative derivation of \ayr{nr/} \fw{nara-} \qq{say, 
speak}: \qq{make sth. said}.

This sentence is also a good illustration for Ayeri's tendency to underspecify
tense. For one, narratives are not marked by the epic past tense, and secondly,
the aspect of actions happening before the time of narration or leading up to it
is only signified by the time adverb \ayr{sris} \fw{sarisa} \qq{previous(ly), 
earlier; former(ly)}.

\divider

\begin{exe}  % SENTENCE 4
    \ex \gll Eng manga yoma -aran eda- mine -ye {} luga bahis -ya kay, nay tong 
	vakas ten puleng -eri, sitang- tong =nama =oy ayon -ang nay envan -ang, 
	nārya nasimaya -ye -ang =hen pandaha -na nay nangān -ena ton naynay. \\
    \mor{AT.INAN} \mor{PROG} exist \mor{-3PL.INAN} this- issue \mor{-PL} 
	\mor{(.T)} for day \mor{-LOC} three, and \mor{3PL.N.A} aware 
	\mor{3PL.INAN.GEN} pain \mor{-INS}, self- \mor{3PL.N.A} =just \mor{=NEG}
	husband \mor{-A} and wife \mor{-A}, but member \mor{-PL} \mor{-A} 
	=all family \mor{-GEN} and household \mor{-GEN} \mor{3PL.N.GEN} as.well \\
    \glt \q{These issues had continued to exist for three days, and they had 
	all been aware of them with pain, not only the husband the wife 
	themselves, but all members of their family and household as well.}
\end{exe}

Words new or modified in this sentence are \ayr{vksF} \fw{vakas} \qq{conscious,
awake}, whose definition should also include \qq{aware}, and \ayr{nNaanF}
\fw{nangān} \qq{household}, which is a renominalization of \ayr{nN} \fw{nanga}
\qq{house}.

\divider

\begin{exe}  % SENTENCE 5
    \ex
    \begin{xlist}
	\ex \label{ex:5a}
	\gll Ang maya -yo nyān {} =hen nanga -ya, ming tenubis -oy -rey, 
	    mitan -tong kadanya. \\
	\mor{AT} feel \mor{-3SG} person \mor{(.T)} =every house \mor{-LOC}, can 
	    make.sense.of \mor{-NEG} \mor{-3SG.INAN}, live \mor{-3PL.N.A} 
	    together.\\
	\glt \q{Every person in the house felt that it couldn't be made sense 
	    of them living together.}
	
	\ex \label{ex:5b}
	\gll Ang eng -yon vih -yam miromān -ye -as keynam {} sisā ang lant 
	    -yo kadanya apin {} kondanga -ya, nasimaya -ye -as pandaha -na nay 
	    nangān -ena Oblonski. \\
	\mor{AT} be.more \mor{3PL.N} share \mor{-PTCP} trait \mor{-PL} \mor{-P} 
	    people \mor{(.T)} \mor{REL-Ø-P} \mor{AT} lead \mor{-3SG.N.A} 
	    together luck \mor{(.T)} inn \mor{-LOC}, member \mor{-PL} \mor{-P} 
	    family \mor{-GEN} and household \mor{-GEN} Oblonski. \\
	\glt \q{People who luck led together in an inn shared more traits than 
	    the members of the family and the household Oblonski.}
    \end{xlist}
\end{exe}

Sentence (\ref{ex:5a}) posed a difficulty in translating \q{there was no sense 
in their living together} \autocite{tolstoy} due to the nominalized VP,
\q{their living together}. I decided to leave the phrase unnominalized in the 
translation, although in principle \ayr{mitMkdnYaanF} \fw{mitankadanyān} 
\qq{living-together} would have been possible. The problem then is, however, 
how to express the genitive part of the original phrase. Fortunately, Ayeri 
allows complement phrases, so \fw{mitantong kadanya} is here literally \qq{that 
they live together}.

Furthermore, I did not want to literally translate \q{there was no sense in …} 
when I discovered that I had earlier coined a verb \ayr{tenubisF/} \fw{tenubis-} 
\qq{make sense of}. Since I wanted to avoid a non-specific \qq{you} here, I 
rendered the subject of this clause as a patient, thus turning the verb into 
a passive form.

Sentence (\ref{ex:5b}) contained the difficulty of a comparative statement that
was beyond a simple \qq{\mor{X} is more \mor{Y} than \mor{Z}}. For literariness, 
I still wanted to use the comparative verb \ayr{ENF/} \fw{eng-} \qq{be more 
\mor{A} than \mor{B}}. I am actually not completely sure if this makes sense,
but I assumed that \ayr{ENF/} \fw{eng-} would be the head in this case and
the phrase headed by \ayr{vihF/} \fw{vih-} \qq{share} would be subordinate to 
it, essentially serving as the quality of \mor{B} that \mor{A} is compared to:

\begin{exe}
    \exi{(ii)} \mor{A} be.more [share traits]_1 than \mor{B} \__1.
\end{exe}

If you look closely at the gloss to (\ref{ex:5b}), you will notice that the NP 
for \mor{B} is marked as a patient, even though according to (ii) it
should be an agent: 

\begin{exe}
    \exi{(iii)}
    \gll nasimaya -ye -as \\
	member \mor{-PL} \mor{-P}. \\
\end{exe}

This is due to an ideosyncrasy in Ayeri by which the standard \mor{B} 
-- as well as a predicative nominal in an equative statement -- is treated like 
a direct object.

The comparative structure is, however, further complicated by the relative 
clause attributed to the agent \ayr{kejnmF(ANF)} \fw{keynam[ang]} \qq{people}. 
Due to information flow and syntactic weight, this constituent moves to the 
right edge of the clause.

\divider

\begin{exe}  % SENTENCE 6
    \ex
    \begin{xlist}
	\ex \gll Ang sara -oy -ye envan {} sangal -as yena, … \\
	\mor{AT} leave \mor{-NEG} \mor{-3SG.F} wife \mor{(.T)} room \mor{-P} 
	    \mor{3SG.F.GEN}, … \\
	\glt \q{The wife did not leave her room, …}
	    
	\ex \gll …, ang manga yoma -oy -ya ayon {} rang -ya ton luga bahis -ya 
	    kay. \\
	…, \mor{AT} \mor{PROG} exist \mor{-NEG} \mor{-3SG.M} husband \mor{(.T)} 
	    home \mor{-LOC} \mor{3PL.N.GEN} for day \mor{-LOC} three. \\
	\glt \q{…, the husband had not been at home for three days.}
    \end{xlist}
\end{exe}

I am a little tempted to insert a conjunction between the two main clauses here,
but I do not know whether it should be \ayr{nj} \fw{nay} \qq{and} or \ayr{ynoymF}
\fw{yanoyam} \qq{because}, since either reading is possible.

\divider

The children ran wild all over the house; the English governess quarreled with the housekeeper, and wrote to a friend asking her to look out for a new situation for her; the man-cook had walked off the day before just at dinner time; the kitchen-maid, and the coachman had given warning.

\begin{exe}  % SENTENCE 7
    \ex
    \begin{xlist}
	\ex …
	\ex …
	\ex …
	\ex …
    \end{xlist}
\end{exe}

%% BIBLIOGRAPHY %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\printbibliography

\end{document}
