\documentclass[12pt,paper=a4]{scrartcl}

% Author, Title, Subtitle etc.
\author{Carsten Becker}
\title{Inunye kora}
\subtitle{Weirdly arpeggiated fishes}
\date{2024/09} % Format: YYYY/MM/DD (rev. N, YYYY/MM/DD)

% Provide running author and date
\makeatletter
\let\runauthor\@author
\let\rundate\@date
\makeatother

% Handle language and quotation marks
\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage{german}
\usepackage{csquotes} % Put quotations in \enquote{}!
\SetBlockEnvironment{quotation}
\renewcommand*{\mkccitation}[1]{ (#1)}
\let\q\textquote

% Quotation style for word definitions
\DeclareQuoteStyle{wdef}
	{\textquoteleft}{\textquoteright}
	{\textquotedblleft}{\textquotedblright}
\newcommand{\wdef}[1]{{\setquotestyle{wdef}\enquote{#1}}}

% Set all margins to 2.54 cm
\usepackage[margin=1in]{geometry}
\widowpenalty10000 % Avoid widows like the plague!
\clubpenalty10000 % Avoid orphans like the plage, too!

% Extended formatting of lists
\usepackage{enumitem}
\newlist{glossdefs}{itemize}{1}
\setlist[glossdefs]{nosep, leftmargin=3em, labelwidth=2.5em, align=left}
\setlist[itemize]{noitemsep}

% Make multiple columns available in single-column document
\usepackage{multicol}

% Make text colors and color names available
\usepackage{xcolor}

% Load font stuff for XeTeX
\usepackage{fontspec}
\usepackage{microtype}

% Set main fonts
\newfontfamily{\Tagati}[
	Renderer=Graphite,
	Scale=1.0,
	BoldFont={* Italic},
	HyphenChar=·,
]{Tagati Book G}

\usepackage{ebgaramond}

\setsansfont[
	Ligatures=TeX,
	Numbers=Lowercase,
	Scale=MatchUppercase,
	BoldFont={Open Sans Condensed Bold},
]{Open Sans Condensed Light}

% Load BibLaTeX (using Biber), configure citation styles
\usepackage[
	style=langsci-unified,
	backend=biber,
	safeinputenc,
	natbib,
]{biblatex}
\addbibresource{bibliography.bib}

% Date etc.
\usepackage[
	datesep={/},
]{datetime2}

% Verse
% \usepackage{verse}

% Tabbing
\usepackage{tabto}

% Ability to include graphics and dealing with footnotes in descriptions
\usepackage{graphicx}
\usepackage[font={small,sf},labelfont={small,sf},format=plain]{caption}
\usepackage{subcaption}
\usepackage{wrapfig}
\setlength{\columnsep}{2\baselineskip}

% General headers and footers
\usepackage{fancyhdr}
\pagestyle{fancy}

\fancyhead[L]{} % empty
\fancyhead[C]{} % empty
\fancyhead[R]{\thepage}

\fancyfoot[L]{} % empty
\fancyfoot[C]{} % empty
\fancyfoot[R]{} % empty

\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

% First page headers and footers are different
\fancypagestyle{firstpage}{
	\fancyhead[L]{\sffamily \footnotesize \textbf{Benung. The Ayeri Language Resource}}
	\fancyhead[C]{} % empty
	\fancyhead[R]{\sffamily \footnotesize \runauthor{} · \rundate{}}
	
	\fancyfoot[L]{} % empty
	\fancyfoot[C]{} % empty
	\fancyfoot[R]{\sffamily \footnotesize 
	\href{https://ayeri.de}{https://ayeri.de} · 
	\href{https://github.com/carbeck/benung-pdfs}{https://github.com/carbeck/benung-pdfs} · 
	\href{https://creativecommons.org/licenses/by-nc-sa/4.0/}{CC~BY-NC-SA~4.0}
	}
	
	\renewcommand{\headrulewidth}{0.5pt}
	\setlength\footskip{0.5in}
}

\usepackage{ifthen}
\ifthenelse{\value{page}=1}{\thispagestyle{firstpage}{\pagestyle{fancy}}}

% Line spacing
\usepackage{setspace}
\onehalfspacing

% Avoid pagebreaks right after sections and subsections
\usepackage{needspace}
\usepackage{etoolbox}
\preto\section{\needspace{6\baselineskip}}
\preto\subsection{\needspace{6\baselineskip}}

% Things for tables
\usepackage{tabularx}
% \usepackage{booktabs}
% \usepackage{rotating}

% Formatting of table of glossing abbreviations from Leipzig package manual
\usepackage[
	acronym,
	nomain,
	nonumberlist,
	nopostdot,
	toc=true,
	xindy={codepage=utf8}, % language=english,
]{glossaries}
\usepackage{glossary-inline}

\newglossarystyle{threecolumn}{%
	\renewenvironment{theglossary}{%
		\begin{multicols}{3}
		\begin{glossdefs}%
	}{%
		\end{glossdefs}%
		\end{multicols}%
	}%
	\renewcommand*{\glossaryheader}{}%
	\renewcommand*{\glsgroupheading}[1]{}%
	\newcommand*{\glossaryentryfield}[5]{%
		\item[\glsentryitem{##1}\glstarget{##1}{##2}]
		% \makefirstuc{##3}\glspostdescription{}
		##3\glspostdescription{}
	}%
	\renewcommand*{\glsgroupskip}{}%
}%

% Formatting of glosses
\usepackage{langsci-gb4e}
\usepackage[glosses]{leipzig}
\renewcommand{\eachwordone}{\itshape}
\renewcommand{\eachwordtwo}{\rule[-.5\baselineskip]{0pt}{0pt}}

\newleipzig{AgtT}{at}{agent topic}
\newleipzig{PatT}{pt}{patient topic}
\newleipzig{DatT}{datt}{dative topic}
\newleipzig{GenT}{gent}{genitive topic}
\newleipzig{LocT}{loct}{locative topic}
\newleipzig{InsT}{inst}{instrumental topic}
\newleipzig{CauT}{caut}{causative topic}
\newleipzig{An}{an}{animate}
\newleipzig{Inan}{inan}{inanimate}
\newleipzig{Hab}{hab}{habitative}
\newleipzig{Sup}{sup}{superlative}
\newleipzig{St}{st}{strong}
\newleipzig{Wk}{wk}{weak}
\newleipzig{Ayr}{ayr}{Ayeri}
\newleipzig{Aff}{aff}{affectionate}

\makeglossaries

% Nicer footnotes
\usepackage[bottom,hang,norule]{footmisc}
\setlength{\footnotesep}{0.75\baselineskip}

% Smaller font in block quotes
\usepackage{relsize}
\AtBeginEnvironment{quote}{\noindent\smaller}
\AtBeginEnvironment{quotation}{\smaller}

% Clickable links in footnotes, TOC, etc.
\usepackage[
% 	xetex,
	bookmarks=true,
	colorlinks=false,
	linktoc=section,
	hidelinks,
	pdfusetitle,
]{hyperref}

% We want URLs to be italic and with regular uppercase numerals
\renewcommand{\UrlFont}{%
	\normalfont%
	\itshape%
	\addfontfeature{RawFeature=-onum}%
}

% In-text references
% cf. https://tex.stackexchange.com/a/139051
% Since German plural formation is not as regular as in English (-e, -en for
% Beispiel), we will define the label as empty
\usepackage[sort&compress,noabbrev]{cleveref}
% \newcommand{\crefrangeconjunction}{--}
\crefname{xnumi}{}{}
\creflabelformat{xnumi}{(#2#1#3)}
\crefname{xnumii}{}{}
\creflabelformat{xnumii}{(#2#1#3)}
\crefname{xnumiii}{}{}
\creflabelformat{xnumiii}{(#2#1#3)}
\crefname{xnumiv}{}{}
\creflabelformat{xnumiv}{(#2#1#3)}
\crefrangeformat{xnumi}{(#3#1#4) to~(#5#2#6)}
\crefrangeformat{xnumii}{(#3#1#4--#5\crefstripprefix{#1}{#2}#6)}
\crefrangemultiformat{xnumii}{(#3\arabic{xnumi}#1#4--#5#2#6)}
{ and~(#3\arabic{xnumi}#1#4--#5#2#6)}{, (#3\arabic{xnumi}#1#4--#5#2#6)}
{ and~(#3\arabic{xnumi}#1#4--#5#2#6)}


% Macros
\newcommand{\fw}[1]{\textit{#1}} % Foreign Word
\newcommand{\tit}[1]{\textit{#1}} % Title of a work
\newcommand{\tsup}[1]{\textsuperscript{#1}} % Superscript
\newcommand{\markyellow}[1]{\colorbox{yellow}{#1}} % Yellow highlighter
\newcommand{\ques}{\textsuperscript{?}} % raised question mark
\newcommand{\zwsp}{\mbox{​}} % Zero-width space (ZWSP)

\newcommand{\ayr}[1]{\zwsp\smash{{\Tagati #1}}} % Plain Ayeri orthography
\newcommand{\rayr}[2]{\zwsp\smash{{\Tagati #1}} \emph{#2}} % Ayeri orthography + *r*omanization
\newcommand{\tayr}[2]{#1 `#2'} % Romanization + *t*ranslation
\newcommand{\xayr}[3]{\zwsp\smash{\Tagati #1} \emph{#2} `#3'} % Ayeri orthography + romanization + translation

\usepackage{suffix}
\WithSuffix\newcommand{\ayr}*[1]{{\Tagati #1}} % Plain Ayeri orthography
\WithSuffix\newcommand{\rayr}*[2]{{\Tagati #1} \emph{#2}} % Ayeri orthography + *r*omanization
\WithSuffix\newcommand{\xayr}*[3]{{\Tagati #1} \emph{#2} `#3'} % Ayeri orthography + romanization + translation

\newenvironment{ayeri}{
	%\hyphenpenalty=10000
	%\hbadness=10000
	\doublespacing
	\begin{multicols}{2}
	\Tagati
}{
	\end{multicols} \par
}

\newenvironment{mytitle}{
	\hfill
	\begin{minipage}{0.667\textwidth}
	\vspace{\baselineskip}
	\begin{center}
		\Large
		\sffamily\bfseries
		\makeatletter
}{
		\makeatother
	\end{center}
	\vspace{1em}
	\end{minipage}
	\hfill
}

%% END OF PREAMBLE %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% MAIN PART %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{mytitle}
	\@title: \@subtitle
\end{mytitle}

At least in my opinion, \citetitle{radiohead:weirdfishes} is the most
fascinating song on
\citeauthor{radiohead:weirdfishes}'s~(\citeyear{radiohead:weirdfishes}) album
\citefield{radiohead:weirdfishes}[booktitle]{booktitle} with its sparkling
string sound. It recently occurred to me that the lyrics are very
straightforward and the spheric soundscape affords one not to be overly strict
with the meter when translating it. All things considered, it seemed worthwhile
to try making an Ayeri adaptation, because translating poetic texts is always a
fun challenge.%
%
	\footnote{This derivative fan work is neither endorsed nor licensed by
	Radiohead, their publisher, or the legal representatives of either of them.
	It neither pursues nor implies any intent of financial gain.}

\begin{quote}
\begin{multicols}{2}
Karonya tasang-vā,\\
tasangya makārya---\\
Sā tila-\\
yang niva.\\
Sinyaley yamanreng\\
perisānena nā?

Nasyoyya depāng-nama.\\
Māy sa kacvāng ay!\\
Sā tila-\\
yang niva\\
vana dimayjyam.			\tab \textit{(kunangreng)}\\
Yam nasyyang vās lito	\tab \textit{(kunangreng)}\\
na Mavay---				\tab \textit{(kunangreng)}\\
nay lesayang.\\
Ang sagonya enya māy,	\tab \textit{(kunangreng)}\\
ang sungya mimānley.	\tab \textit{(kunangreng)}\\
Nay eda-\dots			\tab \textit{(kunangreng)}\\
da-nā.

\columnbreak

Ang tavay kondanley\\
limajyam rina\\
nay kimbisanas\\
inunjyam kora,\\
inunjyam kora,\\
inunjyam kora.

Māy, ang---\\
Ang haray avanya,\\
haray avanya nay\\
nimpyang.\\
Ang---\\
Ang haray avanya\\
haray avanya nay\\
nimpyang.
\end{multicols}

% \begin{multicols}{2}
% \Tagati
% kronFy tsNF/vaa\\
% tsNFy mkaarFy\\
% saa til—\\
% —yNF niv.\\
% sinFylej ymnFreNF\\
% perisaanen naa?

% nsYojy depaaNF/nm.\\
% maaj s ktYFvaaNF Aj!\\
% saa til—\\
% —yNF niv\\
% vn dimjyeymF.			\tab \textit{kunNFreNF}\\
% ymF nsYyNF vaasF lito	\tab \textit{kunNFreNF}\\
% n mvj					\tab \textit{kunNFreNF}\\
% lesyNF.\\
% ANF sgonFy EnFy maaj	\tab \textit{kunNFreNF}\\
% ANF suNFy mimaanFlej.	\tab \textit{kunNFreNF}\\
% nj Ed/—				\tab \textit{kunNFreNF}\\
% d/naa.

% ANF tvj koMdnFlej\\
% limyeymF rin\\
% nj kiMbisnsF\\
% InunFyeymF kor\\
% InunFyeymF kor\\
% InunFyeymF kor

% maaj, ANF\\
% Ang hrj AvnFy\\
% hrj AvnFy nj\\
% niMpFyNF.\\
% ANF\\
% Ang hrj AvanFy\\
% hrj AvnFy nj\\
% niMpFyNF.
% \end{multicols}
\end{quote}

\section{Analysis of the translation}

The song is divided into four parts, of which the first two form a unit in both
text and melody as more and more layers of guitar arpeggi are stacked on top of
each other. Then, after a break, the arpeggi become more muted for the third
part. A concluding and very sober sounding coda follows after an instrumental
section. I will go through my translation part by part, commenting on each
sentence about what stands out to me as worth mentioning regarding grammar,
style, and the process of translation as such.

\subsection{First part}

My translation of the first stanza is given in examples number
\crefrange{ex:1a}{ex:1c}.%
%
	\footnote{Note that where animacy is concerned, I'm only explicitly marking
	inanimate referents in the glosses throughout.}
%
I think that in English, the text corresponding to \cref{ex:1a,ex:1b} can be
interpreted as a single sentence: \blockquote{In the deepest ocean, the bottom
of the sea, your eyes, they turn me.} However, due to the requirements of both
melody and syntax, as well as to keep up the parallelism with \cref{ex:2c}, it
wouldn't have made sense to pull the other half of the statement in
\cref{ex:1b} all the way to the front and have the location adverbials in
\cref{ex:1a} act as modifiers to that, as you'd normally have it in Ayeri. I
thus decided to keep the lines in \cref{ex:1a} at the beginning of the song and
use them to establish context and mood, as in English, though at the expense of
making the second line, \rayr{tsNFy mkaarFy}{tasangya makārya}, an anacoluthon.

\begin{exe}
\ex \label{ex:1a}
	\gll Karonya tasang-vā, tasangya makārya~--- \\
		ocean-\Loc{} deep=\Sup{} abyss-\Loc{} black \\
	\trans `In the deepest ocean, in the black abyss---'
\end{exe}

Consequentially, the text in \cref{ex:1a} doesn't form a complete sentence.
Since \rayr{kronF}{karon} can mean both `sea' and `ocean', I couldn't replicate
the way the English lyrics use these two words for large bodies of water as
basically synonyms. Instead, I played on the fact that \rayr{tsNF}{tasang} can
mean either `deep' or `abyss, chasm' depending on whether it's used as an
adjective or noun, and chose to use \xayr{mkaarFy}{makārya}{dark, black} as an
accompanying adjective accordingly for near-parallelism with \rayr{kronY
tsNF/vaa}{karonya tasang-vā}. The word used for `black' here,
\rayr{mkaarFy}{makārya}, literally means `unwhite'. I don't remember why I
chose to coin the word that way. There is also \xayr{gisu}{gisu}{dark} in the
dictionary, but this is a worse fit considering both the number of syllables of
the original line and assonance in /a/.

The next two lines posed something of a challenge as well, since the English
phrasing \blockquote{Your eyes, they turn me} with what basically amounts to a
topic--comment sentence structure doesn't work this way in Ayeri. Instead,
\cref{ex:1b} can only be translated as a regular statement with no syntactic
break in the middle. The topicalization of \textquote{eyes} can be adapted by
using a regular causative construction, as topic marking is grammaticalized in
Ayeri.

\begin{exe}
\ex \label{ex:1b}
	\gll Sā tilayang niva.\normalfont\footnotemark{} \\
		\CauT{}= change=\Fsg.\Aarg{} gaze \\
	\trans `The gaze makes me change.'
	\footnotetext{\xayr{niv}{niva}{eye}, extended here to mean `gaze'.}
\end{exe}

Since my Ayeri translation of these two lines only amounts to three
words, another question is where to divide the lines. I decided to make the cut
according to the number of syllables, so right inside the word
\rayr{tilyNF}{tilayang} before the heavy syllable
\rayr{/yNF}{-yang}. This slightly changes the distribution of syllables per line
from 2~+~3 in English to 3~+~3 in Ayeri, though it works because
\textquote{your eyes} is rather drawn out in the recording, so another syllable
can be sneaked in easily.

At least to my non-native ears, the phrase \fw{turn somebody} in the English
original of \cref{ex:1b} sounds strange, and looking up the verb in a
dictionary didn't yield any clarification either. So for all intents and
purposes, I suppose that the line can only be understood in the context of its
continuation in \cref{ex:2c}. Additionally, grammatical information about the
owner of the gaze (or eye?), \xayr{vn}{vana}{your}, falls victim to keeping the
line length similar to the English text. The Ayeri verb \rayr{til/}{tila-} can
mean `change' when used intransitively, so the sentence is still more or less
sensible on its own, but the context as to whose gaze and the how or why of
changing is similarly lost.

The next two lines in \cref{ex:1c} can be more easily expressed as a single
statement in Ayeri again, compared to the English \blockquote{Why should I stay
here? Why should I stay?} The 8~+~5 syllables of the English lyrics get
redistributed and shortened by one syllable overall to 6~+~6 in the Ayeri
translation, with the break right between the four words.

\begin{exe}
\ex \label{ex:1c}
	\gll Sinyaley yamanreng perisānena nā? \\
		what-\Parg.\Inan{} reason-\Aarg.\Inan{} hesitation-\Gen{}
		\Fsg.\Gen{} \\
	\trans `What's the reason for my hesitation?'
\end{exe}

In consequence, the urgency of the question evoked by repetition in the English
text doesn't carry over well. I will grant that reading hesitation into the
English lines is my interpretation. Since the song climaxes with thoughts of
escape, though, an expression of ambivalence toward feelings of infatuation at
least doesn't seem far-fetched to me in the context of the song (or Radiohead's
oeuvre as such). Besides this, I decided to reverse the order of the subject
and object noun phrases in \cref{ex:1c} because of the genitive attribute. It
seemed awkward to me to leave \xayr{sinFylej}{sinyaley}{what} dangling at the
end, where it would more usually occur because Ayeri doesn't do question-word
fronting \autocite[153--155]{becker:ayrgram}.

\subsection{Second part}

The next two lines, \blockquote{I'd be crazy not to follow, follow where you
lead} form a unit in English, with the repetition of \textquote{follow} binding
the two together. I made this less explicit in the corresponding Ayeri
translations in \cref{ex:2a,ex:2b}, also because anadiplosis doesn't work well
in Ayeri for syntactic reasons.%
%
	\footnote{I had to look up the term, admittedly.}
%
Literally using \xayr{sdyo}{sadayo}{crazy} in \cref{ex:2a} proved awkward as to
line length, so I rephrased the text as below, though I still had to reach into
my bag of tricks and contracted \rayr{depNNF}{depangang} to
\xayr{depaaNF}{depāng}{fool}. This also eschews the awkwardness of
\rayr{/ANNF}{-angang} with regard to the corresponding English line's rhythm.

\begin{exe}
\ex \label{ex:2a}
	\gll Nasyoyya depāng-nama.\normalfont\footnotemark{} \\
		follow-\Neg-\Tsg.\M{} fool-\Aarg=only \\
	\trans `Only a fool wouldn't follow.'
\end{exe}

I completely rephrased the English text in \cref{ex:2b} as well to yield the
same number of syllables. For rhetoric effect I used a \q{pseudo-passive}
construction with a patient topic and an affective \rayr{maaj}{māy} at the
start \autocite[cf.][361--362]{becker:ayrgram}.

\begin{exe}
\ex \label{ex:2b}
	\gll Māy sa kacvāng ay! \\
		\Aff{} \PatT= draw=\Second.\Aarg{} \Fsg.\Top{} \\
	\trans `Oh how I'm drawn by you!' \\
\end{exe}

At first, I tried \xayr{y nsYFyNF ItiNF vn}{ya nasyyang iting vana}{your path,
I'm following it}, which is slightly closer to the English text, but is longer
by one syllable and thus doesn't fit the melody as well. The rephrased line now
lacks a mention of following, though this motif is at least alluded to by using
\xayr{ktYF/}{kac-}{draw, pull}, and is still explicitly referred to in
\cref{ex:2d} again.

The single sentence in \cref{ex:2c} is actually spread over three lines and
picks up on the text in \cref{ex:1b}, which was noted before to lack context.
In English, the context of \blockquote{your eyes, they turn me} is provided by
\blockquote{turn me into phantoms}. The Ayeri translation follows the English
text very closely here.

\begin{exe}
\ex \label{ex:2c}
	\gll Sā tilayang niva vana dimayjyam. \\
		\CauT= change=\Fsg.\Aarg{} gaze.\Top{} \Second.\Gen{}
		ghost-\Pl-\Dat{} \\
	\trans `Your gaze makes me turn into ghosts.'
\end{exe}

Still, the construction as such is maybe even trickier in relation to its
musical context. Relevant information about the origin of the gaze,
\xayr{vn}{vana}{your}, is stated almost like an afterthought by tearing the
noun phrase apart. This effect is reinforced by musical phrasing, since
\blockquote{turn me into phantoms} and \rayr{vn dimjyeymF}{vana dimayjam},
respectively, forms a unit with the next three lines subsumed in \cref{ex:2d}.
In Ayeri, the feeling of \fw{non sequitur} may thus be even stronger than in
English. Moreover, the haunting repetition of \textquote{way out} as part of
the background vocals starts here, foreshadowing the lines in
\cref{ex:2e,ex:2f}. I chose \xayr{kunNFreNF}{kunangreng}{the door} for the
Ayeri translation in order not to double the verb \xayr{sgonF/}{sagon-}{exit,
quit} with the corresponding noun \xayr{sgonnF}{sagonan}{exit} in \cref{ex:2e}.

The three lines subsumed by the sentence in \cref{ex:2d} again invoke the motif
of following, though the caution introduced in \cref{ex:1c} finds an echo here
in the consequence of falling off the edge of the world after being misled by
infatuation. The translation follows the English text closely, despite there
being no way to mark an action as involving a punctual change of state in the
same way English uses \fw{fall off} as opposed to just \fw{fall}. Moreover, I
decided to use \xayr{mvj}{mavay}{world} as a name to resemble the stress
pattern of the English line. The alternative would have been using
\rayr{mvyen}{mavayena} with penultimate stress and an additional syllable,
which seems worse.

\begin{exe}
\ex \label{ex:2d}
	\gll Yam nasyyang vās lito na Mavay~--- nay lesayang. \\
		\DatT= follow-\Fsg.\Aarg{} \Second.\Parg{} edge.\Top{} \Gen= World
		and fall=\Fsg.\Aarg{} \\
	\trans `To the edge of the Earth is where I'll follow you---and fall.'
\end{exe}

The sentence in \cref{ex:2e} corresponds to the English \blockquote{Yeah,
everybody leaves if they get the chance}, which divides into one clause per
line. The number of syllables in the English text is one less per clause
compared to the Ayeri one: 6~+~5 in English as opposed to 7~+~6 in Ayeri. The
overall stress pattern is similar enough that it works, though. The expression
\xayr{suNF– mimaanFlej}{sung- mimānley}{find the opportunity} strikes me as
rather influenced by German(ic), but offhand, I couldn't think of a different
but succinct way that fits well with the melody.

\begin{exe}
\ex \label{ex:2e}
	\gll Ang sagonya enya māy, ang sungya mimānley. \\
		\AgtT= quit-\Tsg.\M{} everyone.\Top{} \Aff{} \AgtT= find-\Tsg.\M{}
		opportunity-\Parg.\Inan{} \\
	\trans `Yeah, everybody quits, if they find the opportunity.'
\end{exe}

The last two lines in the second part are given in the sentence in
\cref{ex:2f}. The English lyrics are \blockquote{And this way out is my
chance.} Part of the sentence is provided by the background vocals, and I
replicated this peculiarity with \xayr{kunNFreNF}{kunangreng}{door}. I couldn't
find a succinct way to translate \textquote{way out} literally, for instance as
\xayr{ssaanFreNF AgonnFymF}{sasānreng agonanyam}{the way to the outside}, so I
chose metonymy as \emph{my} way out. I also used \xayr{d/naa}{da-nā}{mine}
instead of something more literal like \xayr{mimaanFlej naa}{mimānley nā}{my
chance}, again in order not to exceed the length of the English text.

\begin{exe}
\ex \label{ex:2f}
	\gll Nay eda-kunangreng da-nā. \\
		and this=door-\Aarg.\Inan{} one=\Fsg.\Gen{} \\
	\trans `And this door is mine.'
\end{exe}

From a morphosyntactic point of view, the construction
\xayr{d/naa}{da-nā}{mine} in \cref{ex:2f} is a little peculiar since Ayeri
normally marks predicative complements with the patient case. However, I
determined in \citet[310--311]{becker:ayrgram} that this is not the case with
predicative possessive pronouns. This rules out a form like
*\rayr{d/naalej}{*da-nāley}.

\subsection{Third part}

The third part of the song contains the lines that provide the first part of
the title, \textquote{Weird fishes.} As the sentence in \cref{ex:3} shows, I
extensively rephrased the Ayeri text compared to the original
\blockquote{I get eaten by the worms and weird fishes, picked over by the worms
and weird fishes.}

\begin{exe}
\ex \label{ex:3}
	\setlength{\glossglue}{0pt plus 4pt minus 2pt}
	\gll Ang tavay kondanley limajyam rina\normalfont\footnotemark{}
		nay kimbisanas inunjyam kora. \\
		\AgtT= become=\Fsg.\Top{} food-\Parg.\Inan{} worm-\Pl-\Dat{} slithery
		and prey-\Parg{} fish-\Pl-\Dat{} rare \\
	\trans `I become food for slithery worms and prey for rare fishes.' \\
	\footnotetext{\xayr{rin}{rina}{slippery}, extended here to mean `slithery'.}
\end{exe}

The necessity to rephrase mainly arose from avoiding to coin new words, so I
had to find a different way to express \textquote{picked over} while still
keeping the grammatical structure of the lines in parallel and the meaning
reasonably similar to the English text. The \xayr{limye}{limaye}{worms} thus
gained an adjective \xayr{rin}{rina}{slippery, slithery} in my translation for
parallelism with the adjective \xayr{kor}{kora}{rare} of
\xayr{InunFye}{inunye}{fishes}.

In addition, a literal translation of the first half like \xayr{s koMdFyonF Aj
limye\_aNF nj InunFye\_aNF kepau}{Sa konjon ay limajang nay inunjang kepau}{I
get eaten by the worms and weird fishes} proved awkward in length and rhythm
again. I also didn't like the sound of \xayr{kepau}{kepau}{strange, weird, odd}
with its accent on the final syllable and the rare \emph{au} diphthong. I
replaced it with \rayr{kor}{kora}, referring here to the peculiarity of rarely
observed things, compare Dutch \fw{raar} `strange, odd, weird, unusual'.

As for case marking, I would've rather used the genitive forms of
\xayr{limye}{limaye}{worms} and \xayr{InunFye}{inunye}{fishes}. Adding
\rayr{/n}{-na} to those stems would've made the words longer, though.
Conversely, using the dative case in \rayr{/ymF}{-yam} saves a syllable through
the contraction of \rayr{/yeymF}{-yeyam} to \fw{-jyam}---the changed sound is
not reflected in the Tahano Hikamu spelling, which is more conservative
\autocite[111--112]{becker:ayrgram}. Saving a syllable makes the Ayeri text a
closer fit to the original syllable count of four in the line \textquote{and
weird fishes.} The Ayeri text amounts to five syllables for both \xayr{limyeymF
rin}{limajyam rina}{slithery worms} and \xayr{InunFyeymF kor}{inunjyam
kora}{rare fishes}. The latter is also what is repeated multiple times and even
though the syllable count exceeds that of the corresponding English text, the
melody halfway supports it. English's mere three syllables are hard to match.

\subsection{Fourth part}

The last section also consists of what amounts to one sentence, which is
repeated twice. Its Ayeri adaptation is given in \cref{ex:4b}. The English text
contains a combined anacoluthon and anadiplosis: \textquote{Yeah, I---I hit the
bottom}. Since Ayeri works differently from English and the full clause begins
with the agent-topic marker \rayr{ANF}{ang}, I translated this bit as
\rayr{maaj ANF}{māy ang} and \rayr{ANF}{ang} in the repetition, respectively.

\begin{exe}
\ex \label{ex:4b}
	\gll Māy, ang haray avanya nay nimpyang. \\
		well \AgtT= hit=\Fsg.\Top{} ground-\Loc{} and run=\Fsg.\Aarg{} \\
	\trans `Well, I hit the ground and run.'
\end{exe}

The English text says \textquote{I hit the bottom,} but I was reluctant at
first to copy English and go with \xayr{hrF/}{har-}{hit, punch, smack}. I
considered using the more general \xayr{gFrenF/}{gren-}{reach (for, out)}
instead, though the dictionary---and my vague memory of coining it---implies
that this verb is more about extending your arm than arriving at a destination.
This is what the similar verb \xayr{gireMdF/}{girend-}{arrive} is for, which is
one syllable too long in this context, however. So \rayr{hrF/}{har-} it became
after all, but with the destination \xayr{AvnFy}{avanya}{on the ground} in the
locative case to distinguish it from the victim, in the patient case, of a blow
with one's fist, for example.

Furthermore, my translation only uses \xayr{niMpF/}{nimp-}{run} here instead of
what the dictionary returns for `escape', that is, \rayr{niMpF– mNsr}{nimp-
mangasara}. The adverb \xayr{mNsr}{mangasara}{away (from)} would make the line
too long, but without it, the sentence should still be comprehensible. Also,
I've always tried to avoid introducing prepositional verbs into Ayeri.

\section{Conclusion}

Even though the English lyrics aren't particularly complex and allow some
leeway with regard to line length, the challenge was in finding alternate ways
to phrase lines in order to make the Ayeri text fit. As usual, Ayeri is much
less concise than English because of the former's agglutinative and the
latter's isolating tendencies. I tried not to coin new words as \fw{prankas}
and largely succeeded with that.%
%
	\footnote{\xayr{pFrMk}{pranka}{thing added to make things fit, even the
	score}, a tongue-in-cheek coining in relation to making up a single word
	including /uː/ for the sole purpose of having one at all
	\autocite[cf.][25]{becker:ayrgram}.}
%
Only two words had their definitions slightly extended, and established grammar
had not to be bent much either.

\bigskip
\noindent I guess I should grab my guitar and practice playing the song now.

%% BIBLIOGRAPHY %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \vfill
% \pagebreak

\begingroup\multicolsep=0pt
\printglossary[
	style=threecolumn,
	type=leipzig,
]
\endgroup

%\nocite{*} % returns all entries from the bibliography database
\printbibliography[heading=bibintoc]

\end{document}
